Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
comment duplication of old intro
Updated texts

\textbf{Policy and value function learning}
We learn the policy and value function using TD3 \citep{fujimotoAddressingFunctionApproximation2018}.
However, we follow \citet{yaratsMasteringVisualContinuous2021,zhaoSimplifiedTemporalConsistency2023}
and augment the loss with n-step returns.
The only difference to TD3 is that instead of using the original observations $o_{t}$, we map them through the
encoder $z_{t} = e_{\bar{\theta}}(o_{t})$ and learn the value function and policy in the latent space $z_{t}$.
The value function is updated by minimizing the following objective:
\begin{align} \label{eq:value-loss}
  \mathcal{L}_{Q} &= \E_{\tau \sim \mathcal{D}} \left[ (Q_{\psi_{k}}(e_{\bar{\theta}}(s_{t}), a_{t}) - y)^{2}  \right], \quad  \forall k \in 1, 2 \\
  y &= \sum_{n=0}^{N-1} \gamma^{n} r_{t+n} + \gamma^{n} \min_{k \in \{1,2\}} Q_{\bar{\psi}_{k}}(e_{\bar{\theta}}(s_{t+n}), a_{t+n}), \nonumber
\end{align}
where we use policy smoothing by adding clipped Gaussian noise $\epsilon \sim \mathcal{N} (0,\sigma^{2})$ to the
action $a_{t+n} = \pi_{\eta}(z_{t+n}) + \epsilon$.
Note that we use the target encoder to get the latent states in both the prediction and the target.

@@

\section{Experiments}
\label{sec:experiments}
In this section, we evaluate \our in a wide variety of continuous control tasks from the DeepMind Control (DMC) suite.
We aim to answer the following questions:
\begin{itemize}
    \item How does \our compare to state-of-the-art model-free and model-based algorithms?
    \item Is learning a representation with only latent-state consistency really better than including reward predictions and value predictions?
    \item Can \our's task-agnostic representation aid task-adaptation with its pre-trained representation?
\end{itemize}

@@

\textbf{Normalizing the latent-space improves sample efficiency}
In \cref{fig:normalization_improves_sample_efficiency},

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=1.0\textwidth]{./figs/normalization-ablation.pdf}}
\caption{\textbf{Ablation of normalization}. \our XXXXXX We plot the mean (solid line) and the $95\%$ confidence intervals (shaded) across 5 random seeds, where each seed averages over 10 evaluation episodes.}
\label{fig:normalization-ablation}
\end{center}
\vskip -0.2in
\end{figure*}

@@

\textbf{Reconstruction loss has a detrimental impact}
Recent work \citep{zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022}
has shown that incorporating a reconstruction term into the representation loss can have a negative impact on performance.
To provide a thorough analysis of \our, we include results where we add a reconstruction term to our
representation loss \cref{eq:rep-loss}.
We perform reconstruction at each time step in the horizon.
\cref{fig:multi-task-pretraining} shows that in no environments does reconstruction aid learning.
Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
overall performance.

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=1.0\textwidth]{./figs/reconstruction-loss-ablation.pdf}}
\caption{\textbf{Reconstruction loss has detrimental impact}. \our XXXXXX We plot the mean (solid line) and the $95\%$ confidence intervals (shaded) across 5 random seeds, where each seed averages over 10 evaluation episodes.}
\label{fig:multi-task-pretraining}
\end{center}
\vskip -0.2in
\end{figure*}

>>>>

\textbf{Policy and value function learning}
We learn the policy and value function using TD3, \citep{fujimotoAddressingFunctionApproximation2018})
and follow \cite{yaratsMasteringVisualContinuous2021,zhaoSimplifiedTemporalConsistency2023}
by augmenting the loss with n-step returns.
The only difference to TD3 is that instead of using the original observations $o_{t}$, we map them through the
encoder $z_{t} e_{\theta}(o_{t})$ and learn the value function and policy in the latent space.
The value function is updated by minimizing the following objective:
\begin{align} \label{eq:value-loss}
  \mathcal{L}_{Q} &= \E_{\tau \sim \mathcal{D}} \left[ (Q_{\psi_{k}}(e_{\theta}(s_{t}), a_{t}) - y)^{2}  \right], \quad  \forall k \in 1, 2 \\
  y &= \sum_{n=0}^{N-1} \gamma^{n} r_{t+n} + \gamma^{n} \min_{k \in \{1,2\}} Q_{\bar{\psi}_{k}}(e_{\bar{\theta}}(s_{t+n}), a_{t+n}), \nonumber
\end{align}
where the action is

@@

\section{Experiments}
\label{sec:experiments}
In this section, we evaluate \our in a wide variety of continuous control tasks from the DeepMind Control (DMC) suite \cite{tassa2018deepmind}.
We aim to answer the following questions:
\begin{itemize}
    \item How does \our compare to state-of-the-art model-free and model-based algorithms?
    \item Is learning a representation with only latent-state consistency really better than including reward predictions and value predictions?
    \item Can \our's task-agnostic representation aid task-adaptation with its pre-trained representation?
\end{itemize}

@@

\textbf{Normalizing the latent-space improves sample efficiency}
In \cref{fig:normalization_improves_sample_efficiency}, we compare \our to the model-free baselines Soft-Actor Critic (SAC) \cite{haarnojaSoft2018} and Randomized Ensemble Double Q Learning (REDQ) \cite{chenRandomizedEnsembledDouble2021}, and the model-based TD-MPC \cite{hansenTemporalDifferenceLearning2022} and TCRL \cite{zhaoSimplifiedTemporalConsistency2023} on eight problems from the DeepMind Control Suite. The results show that our method, \our, performs on par with TD-MPC while being significantly simpler and quicker to train, as the loss term is simpler and no planning is needed unlike in TD-MPC. We even outperform TCRL, which is the most similar baseline to our work. We ablate the effect of the FSQ-based latent-space normalization scheme in \cref{fig:normalization-ablation}. The normalization is overall beneficial.

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=1.0\textwidth]{./figs/normalization-ablation.pdf}}
\caption{\textbf{Ablation of normalization}. We analyze the impact of our FSQ-based normalization scheme on the sample-efficiency of \our. We plot the mean (solid line) and the $95\%$ confidence intervals (shaded) across 5 random seeds, where each seed averages over 10 evaluation episodes.}
\label{fig:normalization-ablation}
\end{center}
\vskip -0.2in
\end{figure*}

@@

\textbf{Reconstruction loss has a detrimental impact}
Learning to minimize the observation reconstruction error has been widely applied in model-based reinforcement learning \cite{sutton2018reinforcement, ha2018recurrent, hafnerLearning2019}, and an observation decoder has been a component of the most successful RL algorithms to date \cite{hafner2023mastering}. However, recent work \citep{zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022}
has shown that incorporating a reconstruction term into the representation loss can have a negative impact on performance. They hypothesize that learning to reconstruct the observations is inefficient as they can have irrelevant details and visuals like shading that are uncontrollable by the agent and do not affect the tasks.

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:multi-task-pretraining} shows that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the overall performance, thus lending further support to the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=1.0\textwidth]{./figs/reconstruction-loss-ablation.pdf}}
\caption{\textbf{Reconstruction loss has detrimental impact}. Unlike many methods, such as Dreamer-v3 \cite{hafner2023mastering}, \our neither has an observation decoder nor a reconstruction term in the loss function. We show that adding a reconstruction loss harms the performance of \our across all evaluation environments. We plot the mean (solid line) and the $95\%$ confidence intervals (shaded) across 5 random seeds, where each seed averages over 10 evaluation episodes.}
\label{fig:multi-task-pretraining}
\end{center}
\vskip -0.2in
\end{figure*}
